---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

## Overview

`dta` is a package designed for simple and efficient data wrangling, offering functions to tackle common data manipulation challenges. These include:

-   `dta_bmi()` computes the Body Mass Index (BMI) given weight and height columns in a data frame or tibble.
-   `dta_freq_mrq()` Frequency table for multiple response questions.
-   `dta_mrq()` Split multiple response question column into binary columns.
-   `dta_recode_auto()` Automatically recode categorical variables in a data frame.
-   `dta_recode()` Recode variables in a data frame based on a dictionary.
-   `dta_transpose` Transpose a data frame with specified column as variable names

Among others.

If you are new to `dta`, the best place to start is the [dta package website](https://stemxresearch.github.io/dta).

## Installation

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("stemxresearch/dta")
```

## Usage

```{r, message = FALSE}
library(dplyr)

```

## Getting help

If you encounter a clear bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/stemxresearch/dta/issues).

## Code of conduct

Please note that this project is released with a [Contributor Code of Conduct](https://www.r-project.org/coc-policy.html). By participating in this project you agree to abide by its terms.
