---
title: "Efficient Data Management and Manipulation"
description: >
  This package provides simple and efficient tools for data
  management and manipulation, including functions to clean,
  transform, and format datasets using popular data-handling
  methods. It aims to simplify common data-wrangling tasks.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Efficient Data Management and Manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

```{r setup}
# install.packages(dta) # uncomment if `dta` is not installed
library(dta)
```

This document provides a comprehensive guide to the `dta` package, showcasing
the sample datasets and functions available, along with their usage examples.
The package is designed for efficient data management, transformation, and
exploratory data analysis.

## Sample datasets and dictionaries

The package includes several sample datasets and dictionaries. Below are
descriptions and examples of how to load a few of them.

### Datasets

`data_bmi`: Sample data for body mass index (BMI) calculations.

```{r}
data("data_bmi")
dta_gtable(data_bmi)
```

`data_cancer`: Sample data: Survival of cancer patients in days.

```{r}
data("data_cancer")
dta_gtable(data_cancer)
```

`data_misspelled`: Sample dataset with general information.

```{r}
data("data_misspelled")
dta_gtable(data_misspelled)
```

### Dictionaries

Dictionaries are data values that are applied to update, modify, or standardize
the values in a specified dataset, enabling consistent and efficient data
transformation.

`dict_recode`: A dictionary with variable value mappings with labels and ordered
status.

```{r}
data("dict_recode")
dta_gtable(dict_recode)
```

`dict_misspelled`: A dictionary with variable value mappings with labels and
ordered status.

```{r}
data("dict_misspelled")
dta_gtable(dict_misspelled)
```

## Functions

### Data management and transformation functions

`dta_columns()`

Retrieve column names from a data frame.

```{r}
data(mtcars)
dta_columns(mtcars, .columns = starts_with("c"))
dta_columns(mtcars, .columns = cyl:wt)
dta_columns(mtcars, .columns = c(mpg, hp, vs, gear))
```

`dta_duplicates()`

Find duplicate rows based on specific columns.

```{r}
df <- data.frame(
  id = c(14, 20, 12, 32, 14, 23, 15, 12, 30, 14),
  name = c(
   "Mary", "Mark", "Faith", "David", "Mary", "Daniel", "Christine",
   "Johnson", "Elizabeth", "Mary"
  ),
  age = c(21, 18, 25, 17, 21, 24, 21, 19, 20, 21)
)

result <- dta_duplicates(df)
dta_gtable(result)

result2 <- dta_duplicates(df, .columns = id)
dta_gtable(result2)
```

`dta_label()`

Assign variable labels to data frame or tibble columns.

```{r}
dat <- data.frame(
  age = c(25, 30, 35, 40),
  gender = c("Male", "Female", "Female", "Male"),
  income = c(50000, 60000, 55000, 65000)
)

names <- c("age", "income")
labels <- c("Age in years", "Annual income")

result <- dta_label(
  dat, dict = NULL, .names = names, .labels = labels
)

dta_gtable(result)
```

`dta_recode()`

Recode variables in a data frame based on a dictionary.

```{r}
library(dplyr)
data("data_sample")
glimpse(data_sample) # look at the data type column
data("dict_recode")
dta_gtable(dict_recode)

result <- dta_recode(
  dat = data_sample,
  dict = dict_recode,
  is_force_sequential = TRUE
)
glimpse(result)
```

`dta_replace()`

Correct misspelled data using a dictionary.

```{r}
data("data_misspelled")
dta_gtable(head(data_misspelled))

data("dict_misspelled")
dta_gtable(dict_misspelled)

# Correct the misspelled entries in `dat` using the
# `dict` dictionary

result <- dta_replace(
  dat = data_misspelled, 
  dict = dict_misspelled, 
  .name = variable, 
  .wrong = old, 
  .correct = new
)
dta_gtable(head(result))
```

`dta_transpose()`

Transpose a data frame with specified column as variable names

```{r}
data("data_cancer")
dta_gtable(data_cancer)

df <- dta_transpose(
  dat = data_cancer, .column_to_use_as_variables = cancer_type
)
dta_gtable(df)
```

### Compute and generate functions

`dta_bmi()`

Calculate body mass index (BMI).

```{r}
data("data_bmi")
dta_gtable(data_bmi)

df <- dta_bmi(
  dat = data_bmi,
  .weight = weight,
  .height = height,
  name = body_mass_index,
  digits = 2
)
dta_gtable(df)
```

`dta_bmicat()`

Categorize BMI into weight categories.

```{r}
data("data_bmicat")
dta_gtable(data_bmicat)

# Categorize `bmi` into the standard BMI categories

df <- dta_bmicat(
  dat = data_bmicat,
  .bmi = bmi,
  name = bmi_cat,
  is_extended = FALSE,
  as_factor = TRUE
)
dta_gtable(df)
```

`dta_mrq()`

Split multiple response question column into binary columns.

```{r}
data("data_gadgets")
dat <- data_gadgets
dta_gtable(dat)

# Split `gadgets_owned` column into separate columns.
# The created columns will be logical (i.e. TRUE / FALSE).

df <- dta_mrq(
  dat = dat,
  .column = gadgets_owned,
  delimeter = ", ",
  is_clean_names = TRUE
)
dta_gtable(df)
```

### Exploratory data analysis functions

`dta_freq()`

Get frequency distribution of a specific column.

```{r}
data("data_sample")
tab <- dta_freq(dat = data_sample, .column = region)
dta_gtable(tab)

# Remove the percentage symbol

tab2 <- dta_freq(
  dat = data_sample,
  .column = region,
  is_sorted = TRUE,
  is_decreasing = TRUE,
  add_percent_symbol = FALSE
)
dta_gtable(tab2)
```

`dta_freq_mrq()`

Frequency table for multiple response questions.

```{r}
data("data_sample")

# An example with multiple response variables labelled
# as Yes / No

result <- dta_freq_mrq(
  dat = data_sample,
  .columns = r:excel,
  value = "Yes",
  name = "Programming proficiency"
)
dta_gtable(result)

# Remove the percentage symbol

result2 <- dta_freq_mrq(
  dat = data_sample,
  .columns = r:excel,
  value = "Yes",
  name = "Programming proficiency",
  add_percent_symbol = FALSE
)
dta_gtable(result2)
```

`dta_crosstab()`

Generate cross tabulations with optional percentages and totals.

```{r}
data("data_sample")
df <- data_sample

# Crosstabulation of frequencies (counts)

result <- dta_crosstab(
  dat = df, .row = region, .column = age_group
)
dta_gtable(result)

# Calculate column percentages

result2 <- dta_crosstab(
  dat = df, 
  .row = region,
  .column = age_group,
  cells = "col",
  add_totals = "col"
)
dta_gtable(result2)

# Calculate row percentages

result3 <- dta_crosstab(
  dat = df,
  .row = region,
  .column = age_group,
  cells = "row",
  add_totals = "row"
)
dta_gtable(result3)
```

### Other utilities

`dta_to_numeric()`

Convert numeric strings in a data frame or tibble to numeric numbers.

```{r}
# A data frame with numeric character (a), characters (b) and numeric numbers (c)

df <- data.frame(
  a = c("1", "2", "3"),
  b = c("A", "B", "C"),
  c = c(4, 5, 6)
)
str(df)

df <- dta_to_numeric(df)
str(df)
```

## Conclusion

This document demonstrates the functionality and usage of the `dta` package,
helping users manage, analyze, and transform their data efficiently. For more
details, consult the package documentation.
